### 2019-12-13 ic-journal/multivariate.tex
Up to Section 5.2

Now, the goal can be reduced to several conjunct

@@

There are two ways to represent the \multivariate
substitution $E\{\til P/\til X\}$, that replaces each \hl{free
occurrence} of variables $X_i$ in $E$ with the corresponding
$P_i$: (1) iterately applying the existing \univariate
\HOLinline{\HOLConst{CCS_Subst}}; (2) defining a new \multivariate substitution
function.
% In any case, we (reasonably) assume that all $X_i$ are
%distinct with each other. 
Note that there is the
possibility that $\til P$   syntactically  again contains variables from $\til X$, thus
different orders of iterated substitutions may lead to different results.
% In fact, there are all kinds of issues with
% substitutions and applying them to term-like structures. 
To obtain the
maximal flexibility with a reduced effort, we defined
a \multivariate version of \HOLinline{\HOLConst{CCS_Subst}}
called \HOLinline{\HOLConst{CCS_SUBST}} based HOL's \texttt{finite\_map} theory~\cite{holdesc}.

A finite map of type \HOLinline{\ensuremath{\alpha} \HOLTokenMapto{} \ensuremath{\beta}} is like a function of type
\HOLinline{\ensuremath{\alpha} \HOLTokenTransEnd \ensuremath{\beta}} having only finitely many elements in its
domain. In HOL, an empty finite map is denoted as \HOLinline{\HOLConst{FEMPTY}}.
If \HOLinline{\HOLFreeVar{fm}} is a finite map, its domain (as a set of keys) is denoted as
\HOLinline{\HOLConst{FDOM} \HOLFreeVar{fm}}. Applying \HOLinline{\HOLFreeVar{fm}} on a certain key, say $k$, is denoted
by \HOLinline{\HOLFreeVar{fm} \HOLConst{'} \HOLFreeVar{k}}.

The function \HOLinline{\HOLConst{CCS_SUBST}} takes a finite
map \HOLinline{\HOLFreeVar{fm}} of type \HOLinline{\ensuremath{\alpha} \HOLTokenMapto{} \ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}} and a CCS expression,
and returns another CCS expression in which
all occurrences of variables in \HOLinline{\HOLFreeVar{fm}} domain are substituted with
the corresponding value in \HOLinline{\HOLFreeVar{fm}}. Such a finite map can be built from
the list of variables \HOLinline{\HOLFreeVar{Xs}} and the corresponding targets
\HOLinline{\HOLFreeVar{Ps}} by a helper function \HOLinline{\HOLConst{fromList}} (whose details are
omitted here), while
\HOLinline{\HOLConst{CCS_SUBST}\\\;\ensuremath{(}\HOLConst{fromList}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Ps}\ensuremath{)}\\\;\HOLFreeVar{E}} is
abbreviated as \HOLinline{\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]}\\\;\HOLFreeVar{E}}.
The substitution mechanism is order-independent. For most CCS operators it
just recursively calls itself on subterms. The only
interesting cases are at agent variables and recursion:
\begin{alltt}
   \HOLConst{CCS_SUBST} \HOLFreeVar{fm} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLKeyword{if} \HOLFreeVar{X} \HOLSymConst{\HOLTokenIn{}} \HOLConst{FDOM} \HOLFreeVar{fm} \HOLKeyword{then} \HOLFreeVar{fm} \HOLConst{'} \HOLFreeVar{X} \HOLKeyword{else} \HOLConst{var} \HOLFreeVar{X}\hfill{[CCS_SUBST_var]}

   \HOLConst{CCS_SUBST} \HOLFreeVar{fm} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{)} \HOLTokenDefEquality{}
     \HOLKeyword{if} \HOLFreeVar{X} \HOLSymConst{\HOLTokenIn{}} \HOLConst{FDOM} \HOLFreeVar{fm} \HOLKeyword{then} \HOLConst{rec} \HOLFreeVar{X} \ensuremath{(}\HOLConst{CCS_SUBST} \ensuremath{(}\HOLFreeVar{fm} \HOLSymConst{\ensuremath{\setminus}} \HOLFreeVar{X}\ensuremath{)} \HOLFreeVar{E}\ensuremath{)}
     \HOLKeyword{else} \HOLConst{rec} \HOLFreeVar{X} \ensuremath{(}\HOLConst{CCS_SUBST} \HOLFreeVar{fm} \HOLFreeVar{E}\ensuremath{)}\hfill{[CCS_SUBST_rec]}
\end{alltt}
Variable substitution only occurs on free
variables. In the case of $\recu X E$: if $X$ (a bound variable) is in
the domain of \HOLinline{\HOLFreeVar{fm}}, \HOLinline{\HOLConst{CCS_SUBST}} must continue the
substitution on $E$ using a reduced map~--- without $X$~--- so that all
occurences of $X$ in $E$ are correctly bypassed.
Since \HOLinline{\HOLConst{CCS_SUBST}} only runs
once on each subterm, the possible free variables in  \HOLinline{\HOLFreeVar{Ps}} are never
 substituted. 
% Thus the process is indeed
% order-independent. 
Below we present some key lemmas about
\HOLinline{\HOLConst{CCS_SUBST}}, omitting the proofs:\footnote{Hereafter,
  some new logical constants from HOL's \texttt{pred_set}
and \texttt{list} theories are used (see~\cite{holdesc} for more
details.): \HOLinline{\HOLConst{DISJOINT}} denotes set disjointness;
``\HOLinline{\HOLConst{set}\\\;\HOLFreeVar{Xs}}'' is the set converted from the list \HOLinline{\HOLFreeVar{Xs}};
\HOLinline{\HOLConst{ALL_DISTINCT}} says that
the elements of a list are all distinct; \HOLinline{\HOLConst{MAP}} is
the mapping function from one list to another; \HOLinline{\HOLConst{EVERY}} means
each element of a list satisfies a predicate; \HOLinline{\HOLConst{ZIP}} creates a
new list from two lists of the same length, and each element of the new list is a pair of
elements from the given lists.}

@@

In the above lemmas, the condition
\HOLinline{\HOLConst{DISJOINT}\\\;\ensuremath{(}\HOLConst{BV}\\\;\HOLFreeVar{E}\ensuremath{)}\\\;\ensuremath{(}\HOLConst{set}\\\;\HOLFreeVar{Xs}\ensuremath{)}} (the bound
variables of $E$ and the free substitution variables are disjoint) is
not necessary but makes some proofs easier. 
As these lemmas are
used in the rest work, such disjointness conditions 
are propagated everywhere. 
% (We believe that they can be
% all eliminated in the future.)

@@

As mentioned in
Section~\ref{ss:context}, \univariate contexts and their
guarded companions are defined as predicates over $\lambda$-expressions
of type $CCS\rightarrow CCS$. These predicates, such as
\HOLinline{\HOLConst{CONTEXT}}
(contexts/expressions), \HOLinline{\HOLConst{WG}} (weak guarded contexts),
\HOLinline{\HOLConst{SG}} (guarded contexts) and \HOLinline{\HOLConst{SEQ}} (sequential
contexts), are defined inductively, i.e.~they are built from some ``holes''
in a bottom-up way. For instance, \HOLinline{\HOLConst{WG}\\\;\ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}.\\\;\HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\\\;\HOLSymConst{\ensuremath{\mid}}\\\;\HOLFreeVar{P}\ensuremath{)}}
holds because \HOLinline{\HOLConst{WG}\\\;\ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}.\\\;\HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\ensuremath{)}}
holds as a base case of the inductive definition of \HOLinline{\HOLConst{WG}}:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{a}. \HOLConst{WG} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\ensuremath{)}\hfill{[WG1]}
\end{alltt}
For any agent variable $X$, we have by $\beta$-reduction:
\HOLinline{\ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}.\\\;\HOLFreeVar{l}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{P}\ensuremath{)}\\\;\ensuremath{(}\HOLConst{var}\\\;\HOLFreeVar{X}\ensuremath{)}\\\;\HOLSymConst{\ensuremath{=}}\\\;\HOLFreeVar{l}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var}\\\;\HOLFreeVar{X}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{P}} (or ``$l.X + P$'' in textbook
notation), and the expression is weakly guarded (Def.~\ref{def:guardness}).

>>>>

Now the goal can be reduced to several conjunct

@@

There are two (obvious) ways to \hl{do} % not "represent" but actually "do"
the \multivariate substitution $E\{\til P/\til X\}$, \hl{which} replaces each \hl{free
occurrence} of the variables $X_i$ in $E$ with the corresponding
$P_i$: (1) iterately applying the existing \univariate
\HOLinline{\HOLConst{CCS_Subst}}; (2) defining a new \multivariate substitution
function which substitutes all variables $\til X$ in parallel.
Note that there is the
possibility that $\til P$ syntactically again contains variables from $\til X$, thus
different orders of iterated substitutions may lead to different results.
To obtain the maximal flexibility with reduced efforts, we defined
a \multivariate version of \HOLinline{\HOLConst{CCS_Subst}}
called \HOLinline{\HOLConst{CCS_SUBST}}, based \hl{on} HOL
\texttt{finite\_map} theory~\cite{holdesc}.

A finite map (of the type \HOLinline{\ensuremath{\alpha} \HOLTokenMapto{} \ensuremath{\beta}}) is like a function (of the type
\HOLinline{\ensuremath{\alpha} \HOLTokenTransEnd \ensuremath{\beta}}) having only finitely many elements in its
domain. In HOL, an empty finite map is denoted as \HOLinline{\HOLConst{FEMPTY}}.
If \HOLinline{\HOLFreeVar{fm}} is a finite map, its domain (as a set of keys) is denoted as
\HOLinline{\HOLConst{FDOM} \HOLFreeVar{fm}}. Applying \HOLinline{\HOLFreeVar{fm}} on a certain key, say $k$, is denoted
by \HOLinline{\HOLFreeVar{fm} \HOLConst{'} \HOLFreeVar{k}}.

The function \HOLinline{\HOLConst{CCS_SUBST}} takes a finite
map \HOLinline{\HOLFreeVar{fm}} (of the type \HOLinline{\ensuremath{\alpha} \HOLTokenMapto{} \ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}}) and a CCS expression,
and returns another CCS expression in which
all occurrences of variables in \hl{the finite domain of} \HOLinline{\HOLFreeVar{fm}} are substituted with
the corresponding value in \HOLinline{\HOLFreeVar{fm}}. \hl{Initially,} such a finite map can be built from
a list of variables \HOLinline{\HOLFreeVar{Xs}} and the corresponding substituted terms
\HOLinline{\HOLFreeVar{Ps}} by a helper function \HOLinline{\HOLConst{fromList}} (\hl{defined by us, and} whose details are
omitted here).
\HOLinline{\HOLConst{CCS_SUBST}\\\;\ensuremath{(}\HOLConst{fromList}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Ps}\ensuremath{)}\\\;\HOLFreeVar{E}} is usually
abbreviated as \HOLinline{\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]}\\\;\HOLFreeVar{E}}.
The substitution mechanism is order-independent, \hl{as there is no
ordering in finite maps.} For most CCS operators it
just recursively calls itself on subterms, \hl{while the} only
interesting cases are at agent variables and recursion:
\begin{alltt}
   \HOLConst{CCS_SUBST} \HOLFreeVar{fm} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLKeyword{if} \HOLFreeVar{X} \HOLSymConst{\HOLTokenIn{}} \HOLConst{FDOM} \HOLFreeVar{fm} \HOLKeyword{then} \HOLFreeVar{fm} \HOLConst{'} \HOLFreeVar{X} \HOLKeyword{else} \HOLConst{var} \HOLFreeVar{X}\hfill{[CCS_SUBST_var]}

   \HOLConst{CCS_SUBST} \HOLFreeVar{fm} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{)} \HOLTokenDefEquality{}
     \HOLKeyword{if} \HOLFreeVar{X} \HOLSymConst{\HOLTokenIn{}} \HOLConst{FDOM} \HOLFreeVar{fm} \HOLKeyword{then} \HOLConst{rec} \HOLFreeVar{X} \ensuremath{(}\HOLConst{CCS_SUBST} \ensuremath{(}\HOLFreeVar{fm} \HOLSymConst{\ensuremath{\setminus}} \HOLFreeVar{X}\ensuremath{)} \HOLFreeVar{E}\ensuremath{)}
     \HOLKeyword{else} \HOLConst{rec} \HOLFreeVar{X} \ensuremath{(}\HOLConst{CCS_SUBST} \HOLFreeVar{fm} \HOLFreeVar{E}\ensuremath{)}\hfill{[CCS_SUBST_rec]}
\end{alltt}
Variable substitution only occurs on free
variables. When the term $E$ of ``\HOLinline{\HOLConst{CCS_SUBST}\\\;\HOLFreeVar{fm}\\\;\HOLFreeVar{E}}''
is in the form of $\recu X E'$, if $X$ (a bound variable) is in
the domain of \HOLinline{\HOLFreeVar{fm}}, \HOLinline{\HOLConst{CCS_SUBST}} must continue the
substitution on $E'$ using a reduced map~--- without $X$~--- so that all
occurences of $X$ in $E'$ are correctly bypassed.
Since \HOLinline{\HOLConst{CCS_SUBST}} only runs
once on each subterm, the possible free variables in  \HOLinline{\HOLFreeVar{Ps}} are never
 substituted.

Below we present some key lemmas about
\HOLinline{\HOLConst{CCS_SUBST}}, omitting the proofs:\footnote{Hereafter,
  some new logical constants from HOL's \texttt{pred_set}
and \texttt{list} theories are used (see~\cite{holdesc} for more
details.): \HOLinline{\HOLConst{DISJOINT}} denotes set disjointness;
``\HOLinline{\HOLConst{set}\\\;\HOLFreeVar{Xs}}'' is the set converted from the list \HOLinline{\HOLFreeVar{Xs}};
\HOLinline{\HOLConst{ALL_DISTINCT}} says that
the elements of a list are all distinct; \HOLinline{\HOLConst{MAP}} is
the mapping function from one list to another; \HOLinline{\HOLConst{EVERY}} means
each element of a list satisfies a predicate; \HOLinline{\HOLConst{ZIP}} creates a
new list from two lists of the same length, and each element of the new list is a pair of
elements from the given lists.}

@@

In \hl{some of} the above lemmas, the condition
\HOLinline{\HOLConst{DISJOINT}\\\;\ensuremath{(}\HOLConst{BV}\\\;\HOLFreeVar{E}\ensuremath{)}\\\;\ensuremath{(}\HOLConst{set}\\\;\HOLFreeVar{Xs}\ensuremath{)}} (the bound
variables of $E$ and the free substitution variables are disjoint) is
added to ease the related proofs. This disjointness requirement
is not necessary \hl{and (in theory) can be all eliminated with more efforts.}
\hl{As many of these lemmas are used in proving other theorems, eventually}
the disjointness condition are propagated everywhere.

@@

As mentioned in
Section~\ref{ss:context}, \univariate contexts and their
guarded companions are defined as predicates over $\lambda$-expressions
of the type ``\HOLinline{\ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS} \HOLTokenTransEnd \ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}}''. These predicates, such as
\HOLinline{\HOLConst{CONTEXT}}
(contexts/expressions), \HOLinline{\HOLConst{WG}} (weak guarded contexts),
\HOLinline{\HOLConst{SG}} (guarded contexts) and \HOLinline{\HOLConst{SEQ}} (sequential
contexts), are defined inductively, i.e.~they are built from some ``holes''
in a bottom-up way. For instance, \HOLinline{\HOLConst{WG}\\\;\ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}.\\\;\HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\\\;\HOLSymConst{\ensuremath{\mid}}\\\;\HOLFreeVar{P}\ensuremath{)}}
holds because \HOLinline{\HOLConst{WG}\\\;\ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}.\\\;\HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\ensuremath{)}}
holds as a base case of the inductive definition of \HOLinline{\HOLConst{WG}}:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{a}. \HOLConst{WG} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\ensuremath{)}\hfill{[WG1]}
\end{alltt}
For any agent variable $X$, we have by $\beta$-reduction:
\HOLinline{\ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{t}.\\\;\HOLFreeVar{l}\HOLSymConst{\ensuremath{\ldotp}}\HOLBoundVar{t}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{P}\ensuremath{)}\\\;\ensuremath{(}\HOLConst{var}\\\;\HOLFreeVar{X}\ensuremath{)}\\\;\HOLSymConst{\ensuremath{=}}\\\;\HOLFreeVar{l}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{var}\\\;\HOLFreeVar{X}\\\;\HOLSymConst{\ensuremath{+}}\\\;\HOLFreeVar{P}} (or ``$l.X + P$'' in textbook
notation), and the expression is weakly guarded (Def.~\ref{def:guardness}).
